Class differences in social networks: Evidence from a referral experiment

### 2025-08-05 writing/manuscript.tex
v1.1.6
Chapter 3 to-do: re-read and send

\section*{Abstract}
Economic connectivity, defined as the share of high-SES connections in one's network, is a strong correlate of labor market income. Yet, low-SES individuals are typically at a disadvantage when it comes to knowing the right people. Referral hiring leverages networks and make explicit the role of economic connectivity where taste-based biases could further exacerbate low-SES outcomes. We conduct a field experiment with 734 university students to study the network compositions of different SES groups. We leverage enrollment networks to identify all potential referral candidates and conduct an incentivized referral exercise to reveal SES biases within these choice sets. We find that the university enrollment networks are highly segregated, with low-SES and high-SES individuals having a higher share of same-SES connections in their networks due to program selection (12\% and 31\% respectively). When considering ex post actualized choice sets for the observed referrals, the segregation becomes worse: Low-SES individuals connect with other low-SES individuals at rates 30\% higher than the university share, while high-SES individuals connect with other high-SES individuals at rates 55\% higher than the university share. Yet, we find no bias against low-SES individuals once we account for network structures. We randomly assign half of the participants to a condition where their referral candidate receives a fixed bonus on top of pay-for-performance referral incentives. We find that additional incentives for the referral candidate do not change connection intensity with the referral nor the referral quality. Our findings suggest that systematic segregation patterns in networks that alter choice sets matter more than taste-based SES biases in referrals, and highlight the potential for institutional action in promoting SES diversity. \medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} social capital, social networks, referral hiring, socioeconomic status, field experiment

@@

Referral hiring—the formal or informal process where firms ask workers to recommend qualified candidates for job opportunities—is a common labor market practice that makes differences in social capital evident.\footnote{Referrals solve some frictions in the search and matching process and benefit both job-seekers and employers. As a consequence, referral candidates get hired more often, have lower turnover, and earn higher wages \citep{brown2016informal,dustmann2016referral,friebel2023employee}.} Since referrals originate from the networks of referrers, the composition of referrer networks becomes a crucial channel that propagates inequality. Similar individuals across socio-demographic characteristics form connections at higher rates \citep{mcpherson2001birds}, making across-SES (low-to-high) connections less likely than same-SES connections \citep{chetty2022na}. Referrals will thus reflect similarities in socio-demographic characteristics present in networks even in the absence of biases in the referral procedure—that is, even when referring randomly from one's network according to some productivity criteria.

Yet, experimental evidence shows referrals can be biased even under substantial pay-for-performance incentives beyond what is attributable to differences in network compositions, at least in the case of gender \citep{beaman_job_2018,hederos2025gender}. A similar bias against low-SES individuals may further exacerbate their outcomes. If job information is in the hands of a select few high-SES individuals to whom low-SES individuals already have limited network access due to their lack of economic connectivity, and high-SES referrers are biased against low-SES individuals—referring other high-SES individuals at higher rates than their network composition would suggest—we should expect referral hiring to further disadvantage low-SES individuals.

The empirical question we answer in this paper is whether referrers exhibit bias against low-SES peers after accounting for differences in network SES composition. We also evaluate the causal impact of two different incentive structures on referral behavior.

In this study, we examine inequalities related to SES by curating a university-wide network dataset comprising over 4,500 students for whom classroom interactions are recorded along with individual attributes. We focus on the role of SES in referrals by experimentally investigating whether individuals who are asked to refer a peer tend to refer a same-SES candidate. We also explore potential mechanisms behind referral patterns by randomizing participants into two different incentive structures. To this end, we conducted a lab-in-the-field experiment with 734 students at a Colombian university. We instructed participants to refer a qualified student for tasks similar to the math and reading parts of the national university entry exam (equivalent to the SAT in the US system). To incentivize participants to refer qualified candidates during the experiment, we set earnings to depend on referred candidates' actual university entry exam scores.

Referral hiring in the labor market can range from firm-level formal referral programs asking employees to bring candidates to simply passing on job opportunities between network members \citep{topa2019social}. Since our participants are students at the university and refer based on exam scores, we abstract away from formal referral programs with defined job openings. Our setting instead resembles situations where contacts share opportunities with each other without requiring the referred candidate to take any action and without revealing the referrer's identity. This eliminates reputational concerns since there is no hiring employer. It also establishes a lower bound on the expected reciprocity for the referrer when combined with pay-for-performance incentives \citep{witte2021workers,bandiera2009social}. At the same time, referring based on university entry exam scores is still an objective, widely accepted measure of ability. We show evidence that referrers in our setting not only possess accurate information about these signals but can also screen more productive individuals from their university network.

In a university setting, class attendance provides essential opportunities for face-to-face interaction between students. This is a powerful force that reduces network segregation by providing ample opportunities to meet across SES groups, because of exposure to an equal or higher level of high-SES individuals compared to the general population \citep{chetty2022n}.\footnote{In a different sample from the same university population, \citet{diaz_peer_2025} show this holds true for the highest-SES individuals at this institution, accounting for about 6\% of their sample but less than 5\% of Colombian high-school graduates \citep{fergusson2021desigualdad}.} The very high level of income inequality in Colombia makes SES differences extremely visible in access to tertiary education, where rich and poor typically select into different institutions \citep{jaramillo-echeverri2023}. However, in the particular institutional setting we have chosen for this study, different SES groups mix at this university, allowing us to focus on SES diversity within the institution. At the same time, as students take more classes together, their similarities across all observable characteristics tend to increase \citep{kossinets_origins_2009}. This is an opposite force that drives high- and low-SES networks to segregate. We observe the net effect of these two opposing forces using administrative data and construct class attendance (enrollment) networks for 734 participants based on the number of common courses they have taken together with other students. This allows us to directly identify aggregate characterizations of different SES groups' network compositions as a function of courses taken (e.g., in same-SES share), as well as the individual characteristics of network members who receive referrals among all possible candidates.

We find strong evidence that networks of high- and low-SES participants exhibit same-SES bias. On average, both groups connect with their own SES group at higher rates than would occur randomly given actual group shares at the university (12\% for low-SES and 31\% for high-SES). As students take more courses together within the same program, their networks dwindle in size and become even more homogeneous in SES shares. At 12 courses together (the median number of courses taken together among referrals), the same-SES share increases to 30\% above the university share for low-SES students and 55\% above for high-SES students. We identify selection into academic programs as a key mechanism explaining this phenomenon: The private university where our study took place implements exogenous cost-based program pricing and does not offer SES-based price reductions. This results in programs with very large cost differences within the same university, with some programs costing up to six times the cheapest one. We find that the average yearly fee paid per student increases with SES, and the high-SES share in the most expensive program at the university—medicine—drives a large part of the network segregation across SES groups.

Do segregated networks account for the differences in SES referral rates across SES groups? Same-SES referrals are 17\% more common than referrer networks suggest. Controlling for differences in network compositions, we find that the entirety of the bias is driven by low-SES referrers. We find no bias against low-SES peers beyond what is attributable to differences in network composition. Regardless of SES, participants refer productive individuals, and referred candidates are characterized by a very high number of courses taken together. The latter underlies the impact of program selection on the intensity of social interaction, where participants activate smaller and more homogeneous parts of their networks for making referrals. Our treatment randomized participants across two different incentive schemes by adding a substantial monetary bonus (\$25) for the referred candidate on top of the pay-for-performance incentives. We provide evidence that treatment incentives did not change referral behavior across the same-SES referral rate, the number of courses taken together with the referral candidate, and the candidate's exam scores. We interpret the lack of differences in the number of courses taken together as further evidence that referrals go to strong social ties across both treatments regardless of the incentive structure.\footnote{This follows directly from earlier evidence showing that referrals tend to go to strong ties, i.e., close friends and/or family members \citep{gee2017social,kramarz2014strong,wang2013marriage}.}

Our main empirical contribution to the experimental referral literature is our observation of the entire network that characterizes the referral choice set. Earlier research compares referrals made across different incentive structures and makes inferences about the counterfactual. For example, \citet{beaman_who_2012} compared referrers paid based on their referred candidate's productivity instead of receiving a fixed finder's fee, and \citet{beaman_job_2018} compared referrers who were restricted to refer either a male or female candidate instead of choosing freely. While \citet{pallais_why_2016} recruited a random sample of non-referred workers for comparison with referred ones, none of the previous studies could identify the entire referral choice set and provide a direct comparison to those who were referred by the participants. Observing the entire network allows us to identify biases in referrals in a more natural way, without imposing restrictions on the choice sets. A similar approach to ours is \citet{hederos2025gender}, who elicited friendship networks by asking referrers to name 5 close friends. Their findings suggest only half of those who were referred were from the elicited friendship network, and thus represent an incomplete observation of the entire referral choice set. We take our analysis one step further by requesting referrals from the enrollment network, where we have complete information on every single connection that may or may not receive a referral. This allows us to neatly separate the effect of network composition from any potential biases stemming from the referral procedure itself.

Second, we build upon the earlier work on inequalities in referrals and the role of SES differences. The reliance of labor markets on referrals, coupled with homophily in social networks, can lead to persistent inequalities in wages and employment \citep{montgomery_social_1991,calvo2004effects,bolte_role_2021}. The premise of these models is that referrals exhibit homophily, so that employees are more likely to refer workers of their own race, gender, SES, etc. Supporting evidence shows that low-SES individuals have networks with lower shares of high-SES individuals, which partly explains why they have worse labor market outcomes \citep{stansbury2024class,chetty2022na}. We contribute by separately identifying the role of network homophily (the tendency to connect with similar others) and referral homophily (the tendency to refer similar others). Our results suggest that network homophily, rather than referral homophily, drives SES inequality in our setting.

To our knowledge, \citet{diaz_peer_2025} are the first to study SES biases in referrals, and our study is conceptually the closest to theirs. Drawing from a similar sample at the same institution, \citet{diaz_peer_2025} focus on referrals from first-year students made within mixed-program classrooms and find no evidence for an aggregate bias against low-SES individuals. We also find no aggregate bias against low-SES individuals in referrals beyond what is attributable to differences in network structure. Our setup differs as we sample from students who completed their first year and impose no limits on referring from a classroom. This has several implications: We find that referrals in our setup go to individuals within the same program, and that programs have different SES shares which become even more accentuated as students take more courses together. While networks drive inequality in referral outcomes because of the institutional environment in our sample, we have no reason to believe first-year student networks in \citet{diaz_peer_2025} have similar levels of segregation to begin with. Our findings suggest that implementing more mixed-program courses that allow for across-SES mixing should be a clear policy goal to reduce segregation \citep{alan2023troeasa,rohrer2021po}.

@@

Undergraduate students at the university choose among 32 different academic programs. Students take between 5 to 7 courses per semester, and programs last anywhere between 4 to 12 semesters (2 to 6 years). The majority (64\%) of students are enrolled in the 10 programs described in Appendix Figure \ref{program_share}. Medicine, the largest program by size at the university, lasts for 12 semesters, followed by engineering programs at 10 semesters. Most remaining programs last for about 8 to 10 semesters, with specialized programs for immediate entry into the workforce lasting only 4 semesters.

To map social connections at the university, we construct enrollment networks using administrative data. For each participant, we identify all other undergraduate students with whom they have taken at least one course and create their individual network of university connections. The size of this network depends on how many different students a participant has encountered through coursework, while the intensity of connection is measured by the number of courses taken together. This approach provides a complete picture of each participant's social environment at the university, including detailed characteristics (i.e., SES, academic program, performance) for both the participant and every person in their network.

@@

\begin{align} \label{eq1}
Y_{ij} = \alpha_i + \beta_{1} SES_{ij} + \beta_{2} X_{ij} + \varepsilon_{ij}
\end{align}

@@

The exam consists of five areas (critical reading, mathematics, natural sciences, social sciences, and English). We focus on critical reading and mathematics as these represent two independent and fundemental skills. Critical reading evaluates competencies necessary to understand, interpret, and evaluate texts found in everyday life and broad academic fields (e.g., history). Mathematics assesses students' competency in using undergraduate level mathematical tools (e.g., reasoning in proportions, financial literacy). These together capture performance in comprehending and critically evaluating written material as well as reasoning and problem-solving abilities.

@@

\subsection{Bonus Treatment}
To examine how different incentive structures affect referral selection, we randomly assign a fixed bonus payment for students who get a referral. In the \textbf{Baseline} treatment, only the participants can earn money based on their referral's performance. The \textbf{Bonus} treatment adds an additional fixed payment of \$25 to the peer who gets the referral. This payment is independent of the referral's actual performance (see Table \ref{tab:treatments}).

@@

Table \ref{tab:balance} presents key demographic characteristics and academic performance indicators across treatments (see Appendix Table \ref{tab:selection} for selection). The sample is well-balanced between the \textbf{Baseline} and \textbf{Bonus} conditions and we observe no statistically significant differences in any of the reported variables (all \textit{p} values $> 0.1$). Our sample is characterized by a majority of middle-SES students with about one-tenth of the sample being high-SES students. The test scores and GPA distributions are balanced. On average, participants had taken 3.8 courses together with members of their network, and the average network consisted of 175 peers.

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Balance between treatments}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc c@{}}
  \toprule
   & \multicolumn{1}{c}{\textbf{Baseline}} & \multicolumn{1}{c}{\textbf{Bonus}} & \multicolumn{1}{c}{\textit{\textbf{p}}} \\ 
  \midrule
  Reading score & 64.712 & 65.693 & \multicolumn{1}{c}{0.134} \\
  Math score & 67.366 & 67.597 & \multicolumn{1}{c}{0.780} \\
  GPA & 4.003 & 4.021 & \multicolumn{1}{c}{0.445} \\
  Connections & 173.40 & 176.88 & \multicolumn{1}{c}{0.574} \\
  Courses taken & 3.939 & 3.719 & \multicolumn{1}{c}{0.443} \\
  Low-SES & 0.419 & 0.401 & \multicolumn{1}{c}{0.615} \\
  Middle-SES & 0.492 & 0.506 & \multicolumn{1}{c}{0.714} \\
  High-SES & 0.089 & 0.094 & \multicolumn{1}{c}{0.824} \\
  \midrule
  Observations & 382 & 352 & 734 \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} This table presents balance tests between \textbf{Baseline} and \textbf{Bonus} conditions. $p$-values for binary outcomes are from two-sample tests of proportions; for continuous variables, from two-sample $t$-tests with unequal variances. All reported \textit{p}-values are two-tailed. Reading and math scores are in original scale units out of 100. GPA is grade point average out of 5. Connections refers to the average number of network members. Low-SES, Med-SES, and High-SES indicate SES categories based on strata.
  \end{tablenotes}
  \label{tab:balance}
  \end{threeparttable}
\end{table}

@@

\begin{figure}[H]
  \centering
  \caption{Network size and courses taken together by time spent at the university}\label{connections_class}
  \includegraphics[width=0.6\linewidth]{../figures/connections.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} 
    This figure displays the average number of connections in blue and the average number of courses taken together with connections in grey across semesters completed. Network size (nb. of connections) peaks around 7 semesters before declining as students graduate. Connection intensity (nb. of courses taken) has an increasing trend.
  \end{tablenotes}
\end{figure}

@@

Do referrals across treatments have different outcomes? We compare the performance and the number of courses taken together with the referrer between the \textbf{Baseline} and \textbf{Bonus} treatments in Table \ref{tab:referral_by_treatment}. We find that the the number of courses taken together with referrer, as well as performance measures across Reading, Math, and GPA are similar across treatments. Taken together, the similarities in academic performance and contact intensity suggest these two factors drive referrals regardless of treatment. For this reason, in the remainder of the paper, we report pooled results combining the averages of referral outcomes across treatments.

@@

How do enrollment networks differ across SES groups? We look at how the number of connections (network size) and number of courses taken together (contact intensity) change across SES groups in Figure \ref{connections_by_ses}. Low- and middle-SES students have larger networks but take fewer courses together with network members, while high-SES students have smaller, denser networks. Specifically, both low- and middle-SES students have significantly larger networks than high-SES students ($t = 3.03$, $p = 0.003$ and $t = 2.49$, $p = 0.013$, respectively), but high-SES students take significantly more courses with their network members than both low- ($t = -3.70, p < .001$) and middle-SES ($t = -4.20$, $p < .001$).

@@

What are the diversity-related consequences of SES-driven differences across networks? In terms of network compositions, participants could connect with other SES groups at different rates than would occur randomly depending on their own SES. We illustrate the average network shares conditional on referrer SES for low-SES in Figure \ref{connections_to_low_ses} and for high-SES in Figure \ref{connections_to_high_ses}.\footnote{For sake of brevity we omit middle-SES from this exposition. For the complete relationship, see Appendix Figure \ref{connections_by_ses_app}.} We observe modest deviations from university-wide SES shares in enrollment networks: Low-SES referrers have on average 38.4\% low-SES peers compared to the university average of 34.3\%, while high-SES referrers have 20.4\% high-SES connections compared to the university average of 15.3\%.

@@

\begin{figure}[H]
  \centering
  \caption{SES distribution by program fee}\label{ses_distribution_by_fees}
  \includegraphics[width=0.6\linewidth]{../figures/ses_distribution_by_fees.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure illustrates the distribution of each SES group across programs sorted by fee. the majority of low-SES select into programs with below average cost, while high-SES select into programs with above average cost. Medicine accounts for a third of all high-SES students at this university.
  \end{tablenotes}
\end{figure}

\subsection{Robustness check: Contact intensity and sharing academic programs} \label{Robustness}

Does the number of courses taken together have an independent effect that goes beyond identifying peers in the same academic program? To evaluate this question we leverage our administrative data, and identify peers within the same program: In each individual network we observe the participant-specific academic program for the participant making the referral and alternative-specific academic program for each referral candidate. We add this new variable in our specification and describe our findings in Table \ref{tab:ses-heterogeneity-program}. Being in the same academic program has a substantial positive effect on referral likelihood, with coefficients ranging from 1.257 to 2.198 across all referrer SES groups. This confirms that program affiliation serves as a strong predictor of referral decisions, reflecting increased familiarity. Our comparison of interest is the point estimate for the standardized number of courses taken. Across all three referrer groups, the standardized number of courses taken together maintains its statistical significance after controlling for same program membership. The coefficient magnitudes are expectedly smaller compared to specifications without program controls (ranging from 0.688 to 0.930) as the newly added variable is a moderator: Matching academic programs leads to taking more courses together. The remaining estimates in our model remain robust to the inclusion of the same-program variable with little change in point estimates. The persistence of statistical significance (all $p < 0.001$) suggests that the number of courses taken together has an independent effect on referral decisions. To sum, our measure of contact intensity seems to capture meaningful social interaction patterns that lead to referrals, and go beyond simply identifying matching academic programs.

@@

\section{Conclusion} \label{Conclusion}
In this paper, we study whether SES groups are biased toward one another beyond what is attributable to differences in their networks, and the effects of different incentive structures on referral behavior. Through a lab-in-the-field experiment that leverages enrollment networks at a socially diverse university, we find that the SES biases in referrals originate mostly from network structures, and referrals under performance-pay incentives do not exacerbate existing SES inequalities.

>>>>

\section*{Abstract}
The share of high-SES connections in one's network is a strong correlate of labor market income. We investigate whether differences in high-SES connection shares are exacerbated by SES biases in referral selection. We conduct a lab-in-the-field experiment with 734 Colombian university students making incentivized referrals from their enrollment networks. Randomizing participants between performance-only incentives and performance plus a fixed \$25 bonus for referral recipients, we find no effect of the treatment on referral choices: Referrals consistently go to high-performing peers with whom they take many courses together. We find that while low-SES referrers exhibit strong in-group preferences, middle- and high-SES referrers show no bias toward other groups. Network segregation, driven by students selecting into programs based on their SES, limits cross-SES referral opportunities even without an explicit SES bias. These suggest institutional policies promoting cross-SES contact are key for reducing SES-based inequalities.
\medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} inequality, economic mobility, peer networks, class discrimination, homophily

@@

Research on economic connectivity has focused on two distinct mechanisms that shape cross-SES connections: network composition (who you have the chance to meet inside an institutional environment) versus individual preference (who you choose to connect with among those available). The prevailing hypothesis emerging from the seminal work of \citet{chetty2022n} is that increasing exposure to high-SES individuals under favorable intergroup contact conditions will lead low-SES individuals to connect with them at higher rates. Universities, in this regard, represent a particularly promising setting as they attract higher-than-population shares of high-SES students, and create more opportunities for cross-SES connections. However, credible evidence on biases in individual preferences to connect across SES groups remains limited. One important reason for this gap is the challenge of creating controlled environments that isolate SES biases while accounting for natural variations in social network compositions.

We overcome this challenge through a lab-in-the-field experiment at a Colombian university. Focusing on the role of SES in referral selection, we studied whether individuals who were asked to refer a peer tended to refer a same-SES candidate. We recruited 734 undergraduate students to make incentivized referrals among peers they encountered during their coursework. Referrals were made for the math and critical reading areas of the national university entry exam, and to incentivize performance-based referral selection, participants earned payments up to \$60 per referral based on their nominee's percentile ranking at the university. This setup provided an objective performance benchmark for referrals while SES biases in referral selection could still play a role.

Referrals originated from each participant's unique course enrollment network that we constructed using extensive administrative data. The enrollment network covered each course the referrer had taken with all other undergraduate students at the university (more than 4,500 individuals). It allowed us to observe both characteristics of every potential referral candidate, and the number of courses taken together as a measure of the intensity of interaction between the referrer and the candidate. Referrals from the enrollment networks enabled us to separate network composition (i.e., chance of meeting during coursework and frequency of contact) from SES biases in referral selection (i.e., individual choice in picking a referral). By doing so, we were able to control for naturally varying network compositions with referral candidates at the individual level, and could identify group-level SES biases in referral selection that go beyond mere opportunities to interact at the university.

We randomized participants into two conditions. In the \textbf{Baseline} condition participants made referrals with performance-based incentives only, where their earnings depended on the actual performance of their referrals. In the \textbf{Bonus} condition, participants made referrals with performance-based incentives and an additional fixed bonus (\$25) going to their referral of choice. We designed the \textbf{Bonus} condition to make SES biases in referral selection even more salient. The fixed bonus created incentives to refer peers with high connection intensity (many courses taken together) even if they performed less well, potentially amplifying SES bias since connection intensity and SES background are correlated.

We find that referrals consistently go to higher-performing peers with strong connection intensity (14 vs. 4 courses taken together), regardless of the incentive conditions and the exam area. Pooling across conditions and exam areas, we find that SES bias in referrals is primarily driven by low-SES participants exhibiting in-group preferences: Controlling for network composition, low-SES referrers are 45\% more likely to refer other low-SES peers  and 44\% less likely to refer high-SES relative to middle-SES peers. In contrast, middle- and high-SES referrers show no bias toward other groups.

With 93\% of referrals going to peers within the same academic program with whom referrers have taken many courses together, we find that network composition rather than SES biases explain most observed referral patterns. At the connection intensity where referrals typically occur (median 12 courses together), network segregation becomes stark: low-SES networks contain 44.5\% low-SES peers versus 35\% university-wide (27\% increase), while high-SES networks contain only 15.7\% low-SES peers (55\% decrease from the university average). This segregation means that even without any bias against low-SES peers, high-SES referrers rarely encounter low-SES candidates in their practical choice sets.

Looking for potential mechanisms driving the segregation in enrollment networks, we identify program selection as key. Program fees at the partner university are fixed on a cost basis, and with less than 5\% of undergraduates qualifying for scholarships. The consequence of these institutional policies is that SES groups end up sorting into programs on the basis of their costs, where some programs cost up to six times more on a yearly basis. To sum, even though low-SES are exposed to higher-than-population shares of high-SES students, and high-SES are not biased toward other SES groups, meaningful interaction opportunities are limited due to institutional and organizational factors.

Our findings should be interpreted with some scope conditions. First, our referrals have no direct job consequences, and referring under anonymity presents a lower stake environment with no potential reputational concerns. Nevertheless, we replicate typical findings from earlier referral experiments where performance-based incentives brings in qualified candidates (e.g., \citet{witte2021workers,beaman_who_2012}).

Second, our enrollment networks capture classroom-based interactions and a measure for their intensity rather than broader social networks and close friendships. This approach offers advantages over self-reported friendship network elicitation methods, which suffers from censoring due to size limitations \citep{griffith2022jle}, or social media networks, which may not overlap with actual interactions at the university. We find that courses taken together predicts referral selection beyond program affiliation, suggesting it does capture meaningful variation in connection intensity.

Finally, our setting examines SES bias within a single institution where cross-SES contact is possible, but where the networks of different SES groups are separated due to program selection. The generalizability to contexts with different institutional structures remains an open question for future research.

We contribute to several strands of literature. First, a burgeoning literature studies the effects of SES on labor market outcomes \citep{stansbury2024class,FriedmanLaurison2019GettingOn,laurison2024sf}, with mechanisms including cultural matching and SES-based discrimination in the hiring processes \citep{rivera2012asr,rivera2016asr,galos2024esr,nunez2004class}. We extend this literature by examining the role of referral networks as a specific mechanism through which SES could affect economic opportunities.

A subset of the literature focuses on SES-based differences in social capital and network formation \citep{chetty2022na,engzell2025ajos,michelman2022tqjoe}, with connection intensity \citep{gee2017jole,kramarz2014res,wang2013aejae,sterling2014ms} and homophily \citep{mcpherson2001birds,currarini2009e,jackson2022,bolte2024jle,montgomery_social_1991} driving differences across groups. Based on the pioneering work of \citet{currarini2010pnas}, we contribute by identifying two different types of homophily, and separate whether differential outcomes stem from network composition (who you know) versus taste-based biases (who you choose to interact with). Our findings suggest that rather than focusing on taste-based biases,  implementing mixed-program courses to increase across-SES connection intensity should be a clear policy goal in order to reduce SES-based network segregation (see e.g., for a similar institutional intervention \citet{rohrer2021po}).

Methodologically, we contribute to the literature on job referral experiments. This literature provides causal evidence on why referrals in the labor market are prevalent,\footnote{Referrals solve frictions in the search and matching process and benefit both job-seekers and employers \citep{topa2019social}. Referral candidates tend to get hired more often, have lower turnover, and earn higher wages \citep{obukhova2013ms,brown2016informal,dustmann2016referral}.} and their consequences. Performance-based incentives bring in qualified candidates otherwise not identified by demographics \citep{beaman_who_2012,pallais_why_2016,friebel2023employee,witte2021workers}, at the cost of disadvantaging certain groups \citep{beaman_job_2018,hederos2025gender}. We extend this literature by a simple referral framework where we replicate the earlier findings and causally evaluate the effect of a sizeable monetary bonus for the referral candidate.

@@

Undergraduate students at the university choose among 32 different academic programs. Students take between 5 and 7 courses per semester, and programs last anywhere between 4 and 12 semesters (2 to 6 years). The majority (64\%) of students are enrolled in the 10 programs described in Appendix Figure \ref{program_share}. Medicine, the largest program by size at the university, lasts for 12 semesters, followed by engineering programs at 10 semesters. Most remaining programs last for about 8 to 10 semesters, with specialized programs for immediate entry into the workforce lasting only 4 semesters.

To map social connections at the university, we construct enrollment networks using administrative data. For each participant, we identify all other undergraduate students with whom they have taken at least one course and create their individual network of university connections. The size of this network depends on how many students a participant has encountered through coursework, while the connection intensity is measured by the number of courses taken together. This approach provides a complete picture of each participant's social environment at the university, including detailed characteristics (i.e., SES, academic program, performance) for both the participant and every person in their network.

@@

\begin{align} \label{eq1}
Y_{ij} = \alpha_i + \beta_{1} SES_{ij} + \beta X_{ij} + \varepsilon_{ij}
\end{align}

@@

The exam consists of five areas (critical reading, mathematics, natural sciences, social sciences, and English). We focus on critical reading and mathematics as these represent two independent and fundamental skills. Critical reading evaluates competencies necessary to understand, interpret, and evaluate texts found in everyday life and broad academic fields (e.g., history). Mathematics assesses students' competency in using undergraduate level mathematical tools (e.g., reasoning in proportions, financial literacy). These together capture performance in comprehending and critically evaluating written material as well as reasoning and problem-solving abilities.

@@

\subsection{Bonus Treatment}
To examine how different incentive structures affect referral selection, we randomly assign a fixed bonus payment for students who get a referral. In the \textbf{Baseline} treatment, only the participants can earn money based on their referral's performance. The \textbf{Bonus} treatment adds a fixed payment of \$25 to the peer who gets the referral. This payment is independent of the referral's actual performance (see Table \ref{tab:treatments}).

@@

Table \ref{tab:balance} presents key demographic characteristics and academic performance indicators across treatments (see Appendix Table \ref{tab:selection} for selection). The sample is well-balanced between the \textbf{Baseline} and \textbf{Bonus} conditions, and we observe no statistically significant differences in any of the reported variables (all \textit{p} values $> 0.1$). Our sample is characterized by a majority of middle-SES students with about one-tenth of the sample being high-SES students. The test scores and GPA distributions are balanced. On average, participants had taken 3.8 courses together with members of their network, and the average network consisted of 175 peers.

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Balance between treatments}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcc c@{}}
  \toprule
   & \multicolumn{1}{c}{\textbf{Baseline}} & \multicolumn{1}{c}{\textbf{Bonus}} & \multicolumn{1}{c}{\textit{\textbf{p}}} \\ 
  \midrule
  Reading score & 64.712 & 65.693 & \multicolumn{1}{c}{0.134} \\
  Math score & 67.366 & 67.597 & \multicolumn{1}{c}{0.780} \\
  GPA & 4.003 & 4.021 & \multicolumn{1}{c}{0.445} \\
  Connections & 173.40 & 176.88 & \multicolumn{1}{c}{0.574} \\
  Courses taken & 3.939 & 3.719 & \multicolumn{1}{c}{0.443} \\
  Low-SES & 0.419 & 0.401 & \multicolumn{1}{c}{0.615} \\
  Middle-SES & 0.492 & 0.506 & \multicolumn{1}{c}{0.714} \\
  High-SES & 0.089 & 0.094 & \multicolumn{1}{c}{0.824} \\
  \midrule
  Observations & 382 & 352 & 734 \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} This table presents balance tests between \textbf{Baseline} and \textbf{Bonus} conditions. $p$-values for binary outcomes are from two-sample tests of proportions; for continuous variables, from two-sample $t$-tests with unequal variances. All reported \textit{p}-values are two-tailed. Reading and math scores are in original scale units out of 100. GPA is grade point average out of 5. Connections refer to the average number of network members. Low-SES, Med-SES, and High-SES indicate SES categories based on strata.
  \end{tablenotes}
  \label{tab:balance}
  \end{threeparttable}
\end{table}

@@

\begin{figure}[H]
  \centering
  \caption{Network size and courses taken together by time spent at the university}\label{connections_class}
  \includegraphics[width=0.6\linewidth]{../figures/connections.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} 
    This figure displays the average number of connections in blue and the average number of courses taken together with connections in gray across semesters completed. Network size (nb. of connections) peaks around 7 semesters before declining as students graduate. Connection intensity (nb. of courses taken) has an increasing trend.
  \end{tablenotes}
\end{figure}

@@

Do referrals across treatments have different outcomes? We compare the performance and the number of courses taken together with the referrer between the \textbf{Baseline} and \textbf{Bonus} treatments in Table \ref{tab:referral_by_treatment}. We find that the number of courses taken together with referrer, as well as performance measures across Reading, Math, and GPA are similar across treatments. Taken together, the similarities in academic performance and connection intensity suggest these two factors drive referrals regardless of treatment. For this reason, in the remainder of the paper, we report pooled results combining the averages of referral outcomes across treatments.

@@

How do enrollment networks differ across SES groups? We look at how the number of connections (network size) and number of courses taken together (connection intensity) change across SES groups in Figure \ref{connections_by_ses}. Low- and middle-SES students have larger networks but take fewer courses together with network members, while high-SES students have smaller, denser networks. Specifically, both low- and middle-SES students have significantly larger networks than high-SES students ($t = 3.03$, $p = 0.003$ and $t = 2.49$, $p = 0.013$, respectively), but high-SES students take significantly more courses with their network members than both low- ($t = -3.70, p < .001$) and middle-SES ($t = -4.20$, $p < .001$).

@@

What are the diversity-related consequences of SES-driven differences across networks? In terms of network compositions, participants could connect with other SES groups at different rates than would occur randomly depending on their own SES. Figure \ref{connections_to_low_ses} and Figure \ref{connections_to_high_ses} illustrate the average network shares conditional on referrer SES respectively for low- and high-SES.\footnote{For sake of brevity we omit middle-SES from this exposition. For the complete relationship, see Appendix Figure \ref{connections_by_ses_app}.} We observe modest deviations from university-wide SES shares in enrollment networks: Low-SES referrers have on average 38.4\% low-SES peers compared to the university average of 34.3\%, while high-SES referrers have 20.4\% high-SES connections compared to the university average of 15.3\%.

@@

\begin{figure}[H]
  \centering
  \caption{SES distribution by program fee}\label{ses_distribution_by_fees}
  \includegraphics[width=0.6\linewidth]{../figures/ses_distribution_by_fees.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure illustrates the distribution of each SES group across programs sorted by fee. The majority of low-SES select into programs with below average cost, while high-SES select into programs with above average cost. Medicine accounts for a third of all high-SES students at this university.
  \end{tablenotes}
\end{figure}

\subsection{Robustness check: Connection intensity and sharing academic programs} \label{Robustness}

Does the number of courses taken together have an independent effect that goes beyond identifying peers in the same academic program? To evaluate this question we leverage our administrative data, and identify peers within the same program: In each individual network we observe the participant-specific academic program for the participant making the referral and alternative-specific academic program for each referral candidate. We add this new variable in our specification and describe our findings in Table \ref{tab:ses-heterogeneity-program}. Being in the same academic program has a substantial positive effect on referral likelihood, with coefficients ranging from 1.257 to 2.198 across all referrer SES groups. This confirms that program affiliation serves as a strong predictor of referral decisions, reflecting increased familiarity. Our comparison of interest is the point estimate for the standardized number of courses taken. Across all three referrer groups, the standardized number of courses taken together maintains its statistical significance after controlling for same program membership. The coefficient magnitudes are expectedly smaller compared to specifications without program controls (ranging from 0.688 to 0.930) as the newly added variable is a moderator: Matching academic programs leads to taking more courses together. The remaining estimates in our model remain robust to the inclusion of the same-program variable with little change in point estimates. The persistence of statistical significance (all $p < 0.001$) suggests that the number of courses taken together has an independent effect on referral decisions. To sum, our measure of connection intensity seems to capture meaningful social interaction patterns that lead to referrals, and go beyond simply identifying matching academic programs.

@@

\section{Conclusion} \label{Conclusion}
We investigate whether SES biases in referral selection stem from taste-based preferences in choosing an SES group over others or network segregation. Through a lab-in-the-field experiment with 734 university students making incentivized referrals from complete enrollment networks, we find that institutional factors dominate individual preferences.
